---
title: "Regulation of tissue-specific circadian rhythms in mouse"
shorttitle: "Regulation of tissue-specific circadian rhythms"
author: Jake Yeung
date: "11 May 2015"
output:
  beamer_presentation:
    toc: false
    theme: "Madrid"
    colortheme: "default"
    fonttheme: "default"
    incremental: true
    slide_level: 2
---

# Introduction

## Organisms have adapted to predict periodic signals


```{r load-data, message=FALSE, warning=FALSE, echo=FALSE}
library(plyr)
library(dplyr)
library(ggplot2)
# source("/home/yeung/projects/tissue-specificity/scripts/functions/GetClockGenes.R")
source("/home/yeung/projects/tissue-specificity/scripts/functions/ActivitiesMergedFunctions.R")
source("/home/yeung/projects/tissue-specificity/scripts/functions/SvdFunctions.R")
source("/home/yeung/projects/tissue-specificity/scripts/functions/PlotActivitiesFunctions.R")
source("/home/yeung/projects/tissue-specificity/scripts/functions/PlotGeneAcrossTissues.R")
source("/home/yeung/projects/tissue-specificity/scripts/functions/AlternativeFirstExonsFunctions.R")
load(file = "/home/yeung/projects/tissue-specificity/Robjs/arrayrnaseq.mydat.Robj")
dat.long <- mydat; rm(mydat)
load(file = "/home/yeung/projects/tissue-specificity/Robjs/cov.long.Robj")
load(file = "/home/yeung/projects/tissue-specificity/Robjs/cov.normreads.filt.rhyth.Robj")
load(file = "/home/yeung/projects/tissue-specificity/Robjs/N_and_N.promoter.Robj")
load(file = "/home/yeung/projects/tissue-specificity/Robjs/N.annot.Robj")
# load(file = "/home/yeung/projects/tissue-specificity/Robjs/array.dat.Robj")  # array.dat
load(file = "/home/yeung/projects/tissue-specificity/Robjs/N.long.Robj")
load(file = "/home/yeung/projects/tissue-specificity/Robjs/act.long.Robj")
```

Coordinating internal biological processes with external cues is crucial to organism fitness

\begin{center}
\includegraphics[height=0.4\textheight]{images/circadian-intro-man}
\end{center}

\let\thefootnote\relax\footnotetext{\tiny{Circadian Surprise: How Our Body Clocks Help Shape Our Waistlines. Aubrey and Streeter 2015}}

## Circadian rhythms (~24 hours) underly many physiological processes.

Tissues orchestrate together to carry out function.

\begin{center}
\includegraphics[height=0.5\textheight]{images/circadian-introduction}
\end{center}

\let\thefootnote\relax\footnotetext{\tiny{The Mammalian Circadian Timing System: Synchronization of Peripheral Clocks. Saini et al 2011}}

## How do circadian and tissue-restricted factors interact to produce tissue-specific rhythmic outputs?

\begin{center}
\includegraphics[height=0.6\textheight]{images/circadian-tissue-interaction}
\end{center}

# Methods

## A dataset of temporal gene expression across 12 tissues

\columnsbegin
\column{.7\textwidth}

\begin{center}
\includegraphics[height=0.8\textheight]{images/method-visualization}
\end{center}


\column{.3\textwidth}

>- \scriptsize liver
- \scriptsize kidney
- \scriptsize adrenal gland
- \scriptsize white/brown fat
- \scriptsize muscle
- \scriptsize heart
- \scriptsize aorta
- \scriptsize lung
- \scriptsize brainstem
- \scriptsize cerebellum
- \scriptsize hypothalamus

\columnsend

\let\thefootnote\relax\footnotetext{\tiny{A circadian gene expression atlas in mammals. Zhang et al 2014}}

## Infer activity of transcription factors using a genome-wide linear model

Model expression $E$ in gene $g$, tissue $c$, time $t$ as a function of motif sitecounts $N$ and (unknown) motif activity $A$

$$E_{gct} = \sum_{m} N_{gm} A_{mct} + noise$$

- Expect to capture 5% to 15% of the variance using the linear model
- Goal: to identify regulatory motifs contributing to gene expression in time and across tissues

\let\thefootnote\relax\footnotetext{\tiny{Arnold et al. Bioinformatics, 2012. Balwierz et al. Genome Res, 2014}}

## Infer activity of TFs using a genome-wide linear model
$E_{gct} = \sum_{m} N_{gm} A_{mct} + noise$
\begin{center}
\includegraphics[height=0.7\textheight]{images/matrix-representation}
\end{center}


## Uncover modules of motifs by matrix decomposition

\begin{center}
\includegraphics[height=0.8\textheight]{images/svd-outer-product-mara1}
\end{center}

## Uncover modules of motifs by matrix decomposition

\begin{center}
\includegraphics[height=0.8\textheight]{images/svd-outer-product-mara2}
\end{center}

# Results

## Questions we explored

- Which motifs contribute significantly to the circadian signal across tissues? 
    + One motif module containing core clock regulators explains majority of the circadian signal
- How do regulators interact to generate tissue-restricted circadian rhythms?
    + Alternative promoter usage is a mode in which the cell regulates tissue-restricted circadian rhythms

## First component captures a majority of the circadian signal

```{r screeplot, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
act.fit <- act.long %>%
  group_by(tissue, gene) %>%
  do(.data = ., FitRhythmicWeighted(df = .))

act.fit$pval.adj <- p.adjust(act.fit$pval, method = "BH")

# Filter for rhythmic

pval.adj.cutoff <- 0.0005

rhythmic_genes <- act.fit %>%
  group_by(gene) %>%
  mutate(pval.adj.min = min(pval.adj)) %>%
  filter(pval.adj.min < pval.adj.cutoff)

rhythmic_genes <- unique(rhythmic_genes$gene)

act.filt <- subset(act.long, gene %in% rhythmic_genes)

omega <- 2 * pi / 24

act.complex <- lapply(split(act.filt, act.filt$tissue), function(x){
  ddply(x, .(gene), ProjectToFrequency2, omega = omega, add.tissue = TRUE)
}) %>%
  do.call(rbind, .) %>%
  mutate(magnitude = Mod(exprs.transformed)) %>%
  arrange(desc(magnitude))
# head(act.complex, n = 100)


# SVD on complex matrix ---------------------------------------------------

act.complex.mat <- dcast(data = act.complex, formula = gene ~ tissue, value.var = "exprs.transformed")
rownames(act.complex.mat) <- act.complex.mat[, "gene"]
act.complex.mat <- act.complex.mat[, 2:ncol(act.complex.mat)]

act.svd <- svd(act.complex.mat) 

plot(act.svd$d^2 / sum(act.svd$d ^ 2), main = "Fraction of signal recovered from each component", ylab = "Fraction of signal", xlab = "Components", type = 'o', cex = 2)
```

## First component contains motif module whose phases are tissue-independent  

```{r eigengenes, echo=FALSE, message=FALSE}


# add row and colnames
rownames(act.svd$u) <- rownames(act.complex.mat)
rownames(act.svd$v) <- colnames(act.complex.mat)
rownames(act.svd$v)[which(rownames(act.svd$v) == "Adr")] <- "AdrenalGland"
rownames(act.svd$v)[which(rownames(act.svd$v) == "Adr")] <- "Muscle"

# Plot
theme_set(theme_gray(base_size = 32))
comp <- 1
eigengene <- act.svd$v[, comp]
eigensamp <- act.svd$u[, comp]
eigenval <- act.svd$d[comp]
# rotate to phase of largest magnitude in sample of eigengene
phase.reference <- Arg(eigengene[which(Mod(eigengene) == max(Mod(eigengene)))])
amp.reference <- 2 * Mod(eigengene[which(Mod(eigengene) == max(Mod(eigengene)))])  # 2 times because amplitude should be DOUBLE the half amplitude from Fourier
rotate.factor <- complex(modulus = 1, argument = phase.reference)
# rotate eigengene by -phase ref
eigengene <- eigengene * Conj(rotate.factor) * (1 / amp.reference)
# rotate eigensamp by +phase ref
eigensamp <- eigensamp * rotate.factor * amp.reference

v.plot <- PlotComplex2(eigengene, labels = rownames(act.svd$v), omega = omega, title = paste("Eigenmotif activity (normalized to Liver)")) +
  scale_size(range=c(2, 6)) + xlim(0, 1.1)
print(v.plot)
```

## Motif module containing core clock regulators contributes to a majority of the tissue-wide circadian signal

```{r eigensamples, echo=FALSE, message=FALSE, fig.align='center'}
u.plot <- PlotComplex2(eigensamp, labels = rownames(act.svd$u), omega = omega, title = paste("Eigentissue activity (normalized to Liver)")) +
  scale_size(range=c(0, 10))
print(u.plot)
```

## Questions we explored

>- Which motifs contribute significantly to the circadian signal across tissues? 
    + One motif module containing core clock regulators explains majority of the circadian signal
\pause
- How do regulators interact to generate tissue-restricted circadian rhythms?
    + Alternative promoter usage is a mode in which the cell regulates tissue-restricted circadian rhythms

## Could alternative promoter usage in different tissues produce tissue-restricted rhythms?

\begin{center}
\includegraphics[height=0.6\textheight]{images/alternative-promoter-usage}
\end{center}

## Correlating alternative promoter usage to tissue-restricted circadian rhythms

```{r correlate-alt-promt-usage, echo=FALSE, message=FALSE, fig.align='center'}
# cov.sub <- subset(cov.normreads.filt.rhyth, gene == jgene)
# cov.sub.sum <- cov.sub %>%
#   group_by(transcript, tissue) %>%
#   summarise(mean_reads = mean(norm_reads))

# for sanity checking
jgene <- "Ddc"; jtrans <- "ENSMUST00000066237"
cov.sum <- subset(cov.normreads.filt.rhyth, gene == jgene & transcript == jtrans & !(is.na(rhythmic.or.not)))
ggplot(cov.sum, aes(x = rhythmic.or.not, y = norm_reads)) + 
  geom_boxplot() + 
  ggtitle(paste(jgene, jtrans)) +
  ylab("Fractional first-exon usage") + 
  xlab("Rhythmic or Not")
```

## Alternative promoter usage is associated with tissue-restricted circadian rhythms

15% of 1062 tissue-restricted circadian rhythms are associated with alternative promoter usage

```{r histogram-pvals, warning=FALSE, echo=FALSE, message=FALSE, fig.align='center', fig.height=5}
fit.afe <- cov.normreads.filt.rhyth %>%
  filter(!(is.na(rhythmic.or.not))) %>%
  group_by(transcript, gene) %>%
  do(FitRhythNonRhyth(jdf = .)) %>%
  filter(!is.na(pval))

# summarize by choosing the top for each gene
fit.afe.summary <- fit.afe %>%
  group_by(gene) %>%
  do(SubsetMinPval(jdf = .))
fit.afe.summary$pval.adj <- p.adjust(fit.afe.summary$pval)

# plot histogram of pvalues
qplot(x = fit.afe$pval, geom = "histogram", binwidth = 0.001, xlab = "P-value")
```

## Motif enrichment analysis in alternative promoters

```{r motif-enrichment, warnings=FALSE, echo=FALSE, message=FALSE, fig.align='center'}

jcutoff <- 0.01
genes.hit <- subset(fit.afe.summary, pval.adj < jcutoff)
genes.hit <- genes.hit$gene
max.pval <- max(subset(fit.afe.summary, pval.adj < jcutoff)$pval)  # any pval below this is a hit
fit.afe.hits <- subset(fit.afe, gene %in% genes.hit)

fit.afe.hits$hit.or.not <- factor(mapply(HitOrNot, fit.afe.hits$coef, fit.afe.hits$pval, 
                                         MoreArgs = list(max.pval=max.pval, min.pval=max.pval)), 
                                  levels = c("NotHit", "Neg", "Pos"))

fit.afe.hits.filt <- subset(fit.afe.hits, hit.or.not %in% c("Neg", "Pos"))

# Annotate N.long to hit or not (subset makes things easier)
N.annot.sub <- subset(N.annot, ensemblid %in% fit.afe.hits.filt$transcript)
annot.dic <- setNames(fit.afe.hits.filt$hit.or.not, fit.afe.hits.filt$transcript)
N.annot.sub$hit.or.not <- sapply(N.annot.sub$ensemblid, function(x){
  return(annot.dic[[x]])
})
annot.dic <- setNames(N.annot.sub$hit.or.not, N.annot.sub$saeedid)
N.long.sub <- subset(N.long, promoterid %in% N.annot.sub$saeedid)
N.long.sub$hit.or.not <- sapply(N.long.sub$promoterid, function(x){
  x <- as.character(x)
  return(annot.dic[[x]])
})

# BEGIN TESTING MOTIFS INDIVIDUALLY
jmotif <- "RORA.p2"
ror <- subset(N.long.sub, motif == jmotif)

# ks.ror <- KsTestPosNeg(jdf = ror)
# fit.ror <- FitPosNeg(jdf = ror)
# ks.ror; fit.ror

# ggplot(data = ror, aes(x = hit.or.not, y = sitecount)) + 
#   geom_boxplot() + 
#   ggtitle(jmotif)
ggplot(data = ror, aes(x = sitecount, colour = hit.or.not, fill = hit.or.not)) + 
  geom_density(alpha = 0.5) +
  ggtitle("Sitecounts of ROR motif")
```

# Conclusion

## Conclusions

- Circadian regulators with tissue-independent phase contribute to a majority of the circadian signal 
- Circadian regulators interact with different promoters in a tissue-dependent context

## Thank you

End of my presentation

