---
title: "Regulation of tissue-specific oscillators in mammals"
author: Jake Yeung
date: "11 May 2015"
output:
  beamer_presentation:
    toc: false
    theme: "Madrid"
    colortheme: "default"
    fonttheme: "default"
    incremental: true
    slide_level: 2
---

# Introduction

## Organisms have adapted to predict periodic signals

```{r load-data, message=FALSE, warning=FALSE, echo=FALSE}
library(dplyr)
library(ggplot2)
source("/home/yeung/projects/tissue-specificity/scripts/functions/GetClockGenes.R")
load(file = "/home/yeung/projects/tissue-specificity/Robjs/arrayrnaseq.mydat.Robj")
# dat.long <- mydat; rm(mydat)
# load(file = "/home/yeung/projects/tissue-specificity/Robjs/cov.long.Robj")
# load(file = "/home/yeung/projects/tissue-specificity/Robjs/cov.normreads.filt.rhyth.Robj")
# load(file = "/home/yeung/projects/tissue-specificity/Robjs/N_and_N.promoter.Robj")

```

Coordinating internal biological processes with external cues is crucial to organism fitness

\begin{center}
\includegraphics[height=0.6\textheight]{images/circadian-intro-man}
\end{center}

\let\thefootnote\relax\footnotetext{\tiny{Circadian Surprise: How Our Body Clocks Help Shape Our Waistlines. Aubrey and Streeter 2015}}

## Circadian rhythms (~24 hours) underly many physiological processes.

Tissues orchestrate together to carry out function.

\begin{center}
\includegraphics[height=0.6\textheight]{images/circadian-introduction}
\end{center}

\let\thefootnote\relax\footnotetext{\tiny{The Mammalian Circadian Timing System: Synchronization of Peripheral Clocks. Saini et al 2011}}

## Transcriptional feedback loops drive circadian rhythms

\columnsbegin
\column{.5\textwidth}
\includegraphics[height=0.6\textheight]{images/bmal-per-loop}
\column{.5\textwidth}
\pause

```{r plot-bmal1-per, echo=FALSE, warning=FALSE}
# clockgenes <- GetClockGenes()
# jtissue <- "Mus"
# jgenes <- c("Arntl", "Per2")
# 
# dat.scale <- dat.long %>%
#   group_by(gene, tissue, experiment) %>%
#   filter(gene %in% clockgenes) %>%
#   mutate(scaled = scale(exprs))
# 
# dat.sub <- subset(dat.scale, gene %in% jgenes & experiment == "array" & tissue == jtissue)
# 
# dat.sub$gene <- c("Bmal1", "Period2")
# 
# m <- ggplot(dat.sub, aes(x = time, y = scaled,
#                      group = gene, 
#                      colour = gene)) + 
#   geom_line(size = 2) + 
#   # facet_wrap(~tissue) +
#   ggtitle("Expression of Bmal1 (+ regulator) and Period2 (- regulator) + 
#   ylab(label = "Normalized log expression") +
#   xlab(label = "Circadian time (CT)")
# m
```
\columnsend

\let\thefootnote\relax\footnotetext{\tiny{Molecular architecture of the mammalian circadian clock. Partch et al 2014}}

## How do circadian and tissue-restricted factors interact to produce tissue-specific rhythmic outputs?

\begin{center}
\includegraphics[height=0.6\textheight]{images/circadian-tissue-interaction}
\end{center}

# Methods

## The dataset

# Results

## Results here

Huge result

# Conclusion

## conclusiosn here

Take it home the message